The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor-Couette flow

### 2016-10-05 paper/thingap_submit/thingap_mri_paper_ApJ_submit.tex
thingap submitted materials incl all plots as eps or pdf

\begin{abstract}
The magnetorotational instability (MRI) is a fundamental process of accretion disk physics, but its saturation mechanism remains poorly understood despite considerable theoretical and computational effort. We present a multiple-scales analysis of the non-ideal MRI in the weakly nonlinear regime -- that is, when the the most unstable MRI mode has a growth rate asymptotically approaching zero from above. Here, we develop our theory in a thin-gap, Cartesian channel. Our results confirm the finding by Umurhan et al. (2007) that the perturbation amplitude follows a Ginzburg-Landau equation. We extend these results by performing a detailed force balance for the saturated azimuthal velocity and vertical magnetic field, demonstrating that even when diffusive effects are important, the bulk flow saturates via the combined processes of reducing the background shear and rearranging and strengthening the background vertical magnetic field. We directly simulate the Ginzburg-Landau amplitude evolution for our system and demonstrate the pattern formation our model predicts on long length and time scales. 
\end{abstract}

@@

The weakly nonlinear regime is where the MRI system is nonlinearly unstable to only the most unstable mode of the linear solution. We find the marginal state, where the most unstable linear MRI mode neither grows nor decays, for a set of dimensionless parameters, and then destabilize the system.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../figures/thingap_streamfuncs_velocity_Pm_1E-3.png}
\caption{First order (left), second order (center), and total (right) velocity perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to faster speeds. Colorbar represents azimuthal velocity. We use a constant amplitude $\alpha = \alpha_{saturation}$ and a small parameter $\epsilon = 0.5$.}\label{fig:allorders_velocity}
%\end{figure*}
\vspace*{\floatsep}
%\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/thingap_streamfuncs_Bfield_Pm_1E-3.png}
\caption{As in Figure \ref{fig:allorders_velocity} but for the magnetic field. Streamlines represent magnetic field structure in the vertical-radial plane, where thicker streamlines correspond to higher magnetic field strengths. Colorbar represents azimuthal magnetic field strength.}\label{fig:allorders_Bfield}
\end{figure*}

@@

a real Ginzburg-Landau equation. The saturated solution to Equation \ref{eq:gle} is evidently $\alpha_{saturation} = \pm \sqrt{b/c}$. We plot the first order, second order, and total perturbation structure of the fluid variables in Figures \ref{fig:allorders_velocity} and \ref{fig:allorders_Bfield} with a constant $\alpha_{saturation}$. This is the Ginzburg-Landau equation that was previously found by URM07. Those authors investigated the behavior of this MRI system as a function of $\Pm$. By analyzing the system over several orders of magnitude in $\Pm$, we reproduce the URM07 result that the analytic saturation amplitude scales as $\alpha_{saturation}^2 \propto \Pm^{4/3}$ in a thin-gap geometry when $\Pm \ll 1$. We also find, as in URM07, that the neutral mode amplitude $\beta$ is governed by a simple diffusion equation that decouples from the Ginzburg-Landau equation.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/thingap_saturation_mechanism_plots_velocity_slice60.png}
\caption{Top panel: Mean radial profile for $u_y^{saturated} - u_y^0$, the difference between the saturated azimuthal velocity profile and the initial azimuthal velocity profile (TC flow). Bottom panel: each term in the steady state force balance (Equation \ref{eqset2} with $\partial_t u_{y} = 0$). Saturated quantities are computed with $\alpha_{saturation} = \sqrt{b/c}$ and $\epsilon = 0.5$. The saturated state shows reduced shear in the bulk of the flow, outside of the boundary layers.} \label{fig:sat_comp_vel}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/thingap_saturation_mechanism_plots_Bfield_slice60.png}
\caption{Top panel: Mean radial profile for $B_z^{saturated}$, the saturated vertical magnetic field (black line). $B_z^0 = 1$ is the constant background magnetic field (gray line). Bottom panel: each term in the steady state inductive balance ($\partial_x$ of Equation \ref{eqset4} with $\partial_t B_y = 0$). Saturated quantities are computed with $\alpha_{saturation} = \sqrt{b/c}$ and $\epsilon = 0.5$. The saturated field is pushed to the radial domain boundaries.}\label{fig:sat_comp_Bfield}
\end{figure}

@@

Here our focus is on a physical description of the saturation mechanism. Figure~\ref{fig:sat_comp_vel} shows the saturated mean radial profiles of $u_0 - u_{y} = -q \Omega_0 x - u_{y}$ and each term in the steady state force balance (i.e. Equation~\ref{eqset2} with $\partial_t u_{y} = 0$). In the bulk of the fluid away from the boundary layers, the saturated state shows reduced shear, with little diffusive contribution. This demonstrates that even in a case where diffusive effects are important, the bulk of the fluid saturates by balancing shear and magnetic tension. As discussed at length in \citet{Vasil:2015}, when diffusive effects are not important, it is impossible to rearrange momentum without also rearranging the magnetic field. The \citet{Vasil:2015} model demonstrates saturation without diffusive effects; our results show that outside of the boundary layers, a simultaneous rearrangement of momentum and field occurs. In the boundary layers, the non-linear advection balances viscous dissipation.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/IVP_plot_thingap_AA_phase_2lambdacrit_2.png}
\caption{Evolution of the Ginzburg-Landau amplitude equation (Equation \ref{eq:gle}) on a Fourier $Z$ domain of length $2 \lambda_{crit}$, where $\lambda_{crit} = 2\pi/k_c$ is the critical wavelength of the system. Top left panel shows the evolution of the amplitude observable $\alpha \alpha^*$ on the full $Z$ domain as a function of time $T$. Bottom left panel shows the amplitude $alpha$ at the final timestep shown, where the black line is the real part $\mathrm{Re}\{\alpha(t = t_{final})\}$ and the gray line is the imaginary part $\mathrm{Im}\{\alpha(t = t_{final})\}$. The final amplitude is bounded by the analytic saturation amplitude $\alpha_{saturation} = \pm \sqrt{b/c}$. Top right panel shows the evolution of the phase angle $\phi = \mathrm{arctan} (\mathrm{Im}(\alpha)/\mathrm{Re}(\alpha))$ on the same domain. Bottom panel shows the phase angle as a function of $Z$ for the final timestep. Note that the phase angle is wrapped on a $2\pi$ domain, such that $\pi$ = $-\pi$, as indicated by the circular colorbar.}\label{fig:IVP_2lambdacrit}
%\end{figure*}

%\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/IVP_plot_thingap_AA_phase_2lambdacrit_10.png}%{allorders_velocity.eps}
\caption{As in Figure \ref{fig:IVP_2lambdacrit} but for a $Z$ domain of length $10 \lambda_{crit}$.}\label{fig:IVP_10lambdacrit}
\end{figure*}

@@

In this paper we construct a weakly nonlinear analysis of the MRI using multiple-scales analysis, leading to a real Ginzburg-Landau equation for the nonlinear amplitude, confirming the previous results of \citet{Umurhan:2007hs}. We also confirm their results for the scaling of the analytic saturation amplitude with $\Pm$. We extend their results by constructing a detailed force and inductive balance for the saturated $u_y$ and $B_z$ components. In doing so, we find the saturated state is a complex balance in which reduction of shear and amplification and redistribution of $B_z$ combine to saturate the instability. We also perform direct numerical simulations of the amplitude equation. We demonstrate that complex patterns can organize the flow on long length scales $Z$, though the maximum magnitude of the amplitude $\alpha$ is well predicted by the steady state solution. In Paper II, we make use of the techniques developed here to extend the weakly nonlinear analysis of the MRI to a full cylindrical geometry appropriate for a Taylor-Couette experiment.

>>>>

\begin{abstract}
The magnetorotational instability (MRI) is a fundamental process of accretion disk physics, but its saturation mechanism remains poorly understood despite considerable theoretical and computational effort. We present a multiple scales analysis of the non-ideal MRI in the weakly nonlinear regime -- that is, when the most unstable MRI mode has a growth rate asymptotically approaching zero from above. Here, we develop our theory in a thin-gap, Cartesian channel. Our results confirm the finding by Umurhan et al. (2007) that the perturbation amplitude follows a Ginzburg-Landau equation. We extend these results by performing a detailed force balance for the saturated azimuthal velocity and vertical magnetic field, demonstrating that even when diffusive effects are important, the bulk flow saturates via the combined processes of reducing the background shear and rearranging and strengthening the background vertical magnetic field. We directly simulate the Ginzburg-Landau amplitude evolution for our system and demonstrate the pattern formation our model predicts on long length and time scales. 
\end{abstract}

@@

The weakly nonlinear regime is where the MRI system is nonlinearly unstable to only the most unstable mode of the linear solution. We find the marginal state, where the most unstable linear MRI mode neither grows nor decays, for a set of dimensionless parameters, and then destabilize the system. We examine the system for fiducial parameters comparable to URM07, namely $\Pm = 1.0 \times 10^{-3}$, $\beta = 25$, $q = 1.5$. The system is marginal for a critical wavenumber $k_c = 0.75$ and a critical magnetic Reynolds number $\reym_c = 4.9$.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{thingap_streamfuncs_velocity_Pm_1E-3.png}
\caption{First order (left), second order (center), and total (right) velocity perturbations. Streamlines represent velocity in the vertical-radial plane, where thicker streamlines correspond to faster speeds. Colorbar represents azimuthal velocity. We use a constant amplitude $\alpha = \alpha_{saturation}$ and a small parameter $\epsilon = 0.5$.}\label{fig:allorders_velocity}
%\end{figure*}
\vspace*{\floatsep}
%\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{thingap_streamfuncs_Bfield_Pm_1E-3.png}
\caption{As in Figure \ref{fig:allorders_velocity} but for the magnetic field. Streamlines represent magnetic field structure in the vertical-radial plane, where thicker streamlines correspond to higher magnetic field strengths. Colorbar represents azimuthal magnetic field strength.}\label{fig:allorders_Bfield}
\end{figure*}

@@

a real Ginzburg-Landau equation. The saturated solution to Equation \ref{eq:gle} is evidently $\alpha_{saturation} = \pm \sqrt{b/c}$. We plot the first order, second order, and total perturbation structure of the fluid variables in Figures \ref{fig:allorders_velocity} and \ref{fig:allorders_Bfield} with a constant $\alpha_{saturation}$. This is the Ginzburg-Landau equation that was previously found by URM07. Those authors investigated the behavior of this MRI system as a function of $\Pm$. By analyzing the system over several orders of magnitude in $\Pm$, we reproduce the URM07 result that the analytic saturation amplitude scales as $\alpha_{saturation}^2 \propto \Pm^{4/3}$ in a thin-gap geometry when $\Pm \ll 1$. We also find, as in URM07, that the neutral mode amplitude $\beta(T, Z)$ is governed by a simple diffusion equation that decouples from the Ginzburg-Landau equation.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{thingap_saturation_mechanism_plots_velocity_slice60.pdf}
\caption{Top panel: radial profile for $u_y^{saturated} - u_y^0$, the difference between the saturated azimuthal velocity profile and the initial azimuthal velocity profile (TC flow). Bottom panel: each term in the steady state force balance (Equation \ref{eqset2} with $\partial_t u_{y} = 0$). Saturated quantities are computed with $\alpha_{saturation} = \sqrt{b/c}$ and $\epsilon = 0.5$. The saturated state shows reduced shear in the bulk of the flow, outside of the boundary layers.} \label{fig:sat_comp_vel}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{thingap_saturation_mechanism_plots_Bfield_slice60.pdf}
\caption{Top panel: radial profile for $B_z^{saturated}$, the saturated vertical magnetic field (black line). $B_z^0 = 1$ is the constant background magnetic field (gray line). Bottom panel: each term in the steady state inductive balance ($\partial_x$ of Equation \ref{eqset4} with $\partial_t B_y = 0$). Saturated quantities are computed with $\alpha_{saturation} = \sqrt{b/c}$ and $\epsilon = 0.5$. The saturated field is pushed to the radial domain boundaries.}\label{fig:sat_comp_Bfield}
\end{figure}

@@

Here our focus is on a physical description of the saturation mechanism. Figure~\ref{fig:sat_comp_vel} shows saturated radial profiles of $u_0 - u_{y} = -q \Omega_0 x - u_{y}$ and each term in the steady state force balance (i.e. Equation~\ref{eqset2} with $\partial_t u_{y} = 0$). In the bulk of the fluid away from the boundary layers, the saturated state shows reduced shear, with little diffusive contribution. This demonstrates that even in a case where diffusive effects are important, the bulk of the fluid saturates by balancing shear and magnetic tension. As discussed at length in \citet{Vasil:2015}, when diffusive effects are not important, it is impossible to rearrange momentum without also rearranging the magnetic field. The \citet{Vasil:2015} model demonstrates saturation without diffusive effects; our results show that outside of the boundary layers, a simultaneous rearrangement of momentum and field occurs. In the boundary layers, the non-linear advection balances viscous dissipation.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{IVP_plot_thingap_AA_phase_2lambdacrit_2.pdf}
\caption{Evolution of the Ginzburg-Landau amplitude equation (Equation \ref{eq:gle}) on a Fourier $Z$ domain of length $2 \lambda_{crit}$, where $\lambda_{crit} = 2\pi/k_c$ is the critical wavelength of the system. Top left panel shows the evolution of the amplitude observable $\alpha \alpha^*$ on the full $Z$ domain as a function of time $T$. Bottom left panel shows the amplitude $alpha$ at the final timestep shown, where the black line is the real part $\mathrm{Re}\{\alpha(t = t_{final})\}$ and the gray line is the imaginary part $\mathrm{Im}\{\alpha(t = t_{final})\}$. The final amplitude is bounded by the analytic saturation amplitude $\alpha_{saturation} = \pm \sqrt{b/c}$. Top right panel shows the evolution of the phase angle $\phi = \mathrm{arctan} (\mathrm{Im}(\alpha)/\mathrm{Re}(\alpha))$ on the same domain. Bottom panel shows the phase angle as a function of $Z$ for the final timestep. Note that the phase angle is wrapped on a $2\pi$ domain, such that $\pi$ = $-\pi$, as indicated by the circular colorbar.}\label{fig:IVP_2lambdacrit}
%\end{figure*}

%\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{IVP_plot_thingap_AA_phase_2lambdacrit_10.pdf}%{allorders_velocity.eps}
\caption{As in Figure \ref{fig:IVP_2lambdacrit} but for a $Z$ domain of length $10 \lambda_{crit}$.}\label{fig:IVP_10lambdacrit}
\end{figure*}

@@

In this paper we construct a weakly nonlinear analysis of the MRI using multiple scales analysis, leading to a real Ginzburg-Landau equation for the nonlinear amplitude, confirming the previous results of \citet{Umurhan:2007hs}. We also confirm their results for the scaling of the analytic saturation amplitude with $\Pm$. We extend their results by constructing a detailed force and inductive balance for the saturated $u_y$ and $B_z$ components. In doing so, we find the saturated state is a complex balance in which reduction of shear and amplification and redistribution of $B_z$ combine to saturate the instability. We also perform direct numerical simulations of the amplitude equation. We demonstrate that complex patterns can organize the flow on long length scales $Z$, though the maximum magnitude of the amplitude $\alpha$ is well predicted by the steady state solution. In Paper II, we make use of the techniques developed here to extend the weakly nonlinear analysis of the MRI to a full cylindrical geometry appropriate for a Taylor-Couette experiment.
